---
title: "What Influences the Decision to Have or Not Have Children?"
author: "Anisha Gondesi, Mikayla Lamping, Michaela Suski"
date: "5/8/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(readr)
library(tidyverse)
library(patchwork)
library(tidymodels)
library(rgeos)
library(rworldmap)
library(rworldxtra)
library(tibble)
library(ggmap)
library(spData)
library(sf)
library(ggrepel)
```


There are many reasons why women all over the world choose not to become pregnant and have children.  Across different geographies, populations, and cultures, these reasons may be financial, social, physical or otherwise.  This report aims to help visualize the reasons why women may not want to have children or become pregnant.

\bigskip

```{r, warning = FALSE, message = FALSE}
mortality <- read_csv("global_mortality.csv") %>%
  filter(year == 2015) %>%
  mutate(maternal = `Maternal deaths (%)`,
         neonatal = `Neonatal deaths (%)`,
         `Maternal deaths (%)` = NULL,
         `Neonatal deaths (%)` = NULL) %>%
  select(c(country, country_code, maternal, neonatal)) %>%
  drop_na()
  
haq <- read_csv("haq_index.csv") %>%
  filter(Year == 2015) %>%
  mutate(country = Entity,
         country_code = Code,
         haq = `HAQ Index (IHME (2017))`,
         Entity = NULL,
         `HAQ Index (IHME (2017))` = NULL) %>%
  select(c(country,country_code, haq))
```

```{r, warning = FALSE, message = FALSE}
countries_ <- intersect(mortality$country, haq$country)

mortality <- mortality %>%
  filter(country %in% countries_)
haq <- haq %>%
  filter(country %in% countries_)

global_ <- inner_join(mortality, haq)
```

```{r fig.height = 8, fig.width = 15, warning = FALSE, message = FALSE}
library(maps)
world_map <- map_data(map = "world") %>% 
  filter(region != "Antarctica")

global_$country[182] = "USA"
global_$country[46] = "Democratic Republic of the Congo"
global_$country[42] = "Ivory Coast"
global_$country[40] = "Republic of Congo"
```

```{r fig.height = 10, fig.width = 12, warning = FALSE, message = FALSE}
p1 <- ggplot() +
  geom_map(data = global_, aes(map_id = country, fill = maternal), map = world_map) +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), colour = 'white', fill = NA, size = 0.2) +
  scale_fill_distiller(palette = "RdPu", direction = 1) +
  labs(fill = "Maternal Mortality Rate") +
  theme_void() 
```

```{r fig.height = 5, fig.width = 7, warning = FALSE, message = FALSE}
p2 <- ggplot() +
  geom_map(data = global_, aes(map_id = country, fill = haq), map = world_map) +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), colour = 'white', fill = NA, size = 0.2) +
  scale_fill_distiller(palette = "RdPu", direction = 1) +
  labs(fill = "HAQ Index", 
       caption = "*HAQ stands for Healthcare Access and Quality") +
  theme_void()

(p1 / p2) + plot_annotation(title = "Maternal Mortality Rate and HAQ* Index by Country") &
  theme(legend.text=element_text(size=8),
        legend.title=element_text(size=10))
```



## Access to Quality Healthcare

Though maternal health outcomes are improving by the day, huge disparity still exists in different areas of the world.  Lower- and middle-income countries do not have the same access to quality healthcare facilities or medical professionals as upper-income countries.

There is an inverse relationship between the colors of the maps above.  This means that countries with lower Healthcare Access and Quality (HAQ) Indexes typically also have higher maternal mortality rates.  Clearly, access to and quality of healthcare is essential for healthy birth outcomes.

The relationship between HAQ Index and Maternal Mortality Rate is illustrated in the following scatterplot as well.  Additionally, the relationship between HAQ Index and Birth Attendant Skill is explored.  Birth Attendant Skill refers to the percentage of births that were attended by skilled medical professionals.

\bigskip

```{r, warning = FALSE, message = FALSE}
skilled <- read_csv("birthAttendedBySkilledPersonal.csv") %>% 
  mutate(country = Location,
         skilled_attendant = `First Tooltip`,
         Location = NULL,
         `First Tooltip` = NULL) %>%
  group_by(country) %>% 
  filter(row_number() == 1) %>%
  select(c(country, skilled_attendant))

skilled$country[20] = "Bolivia"
skilled$country[46] = "North Korea"
skilled$country[47] = "Democratic Republic of Congo"
skilled$country[80] = "Iran"
skilled$country[92] = "Laos"
skilled$country[136] = "South Korea"
skilled$country[140] = "Russia"
skilled$country[163] = "Syria"
skilled$country[178] = "United States"
skilled$country[182] = "Venezuela"
skilled$country[183] = "Vietnam"
```

```{r, warning = FALSE, message = FALSE}
continents <- read_csv("countries-continents.csv") %>%
  mutate(country = Country,
         continent = Continent) %>%
  select(country, continent)
cont_countries <- intersect(countries_, continents$country)

skilled_ <- skilled %>%
  filter(country %in% cont_countries)

global <- inner_join(global_, skilled_)
global <- inner_join(global, continents)
```

```{r fig.height = 8, fig.width = 8, warning = FALSE, message = FALSE}
p3 <- suppressWarnings(ggplot(global) +
  geom_point(aes(haq, maternal, col = continent), size = 1.8) +
  geom_text_repel(aes(haq, maternal, 
                      label = ifelse(maternal > 1.5 & haq < 55, country, ''),
                      segment.size = 0.2), 
                  size = 4, max.overlaps = 27) +
  labs(title = "Maternal Mortality Rate vs. Heathcare Access and Quality Index", 
       x = "HAQ Index", 
       y = "Maternal Mortality Rate") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()
)
```

```{r fig.height = 10, fig.width = 10, warning = FALSE, message = FALSE}
p4 <- suppressWarnings(ggplot(global) +
  geom_point(aes(haq, skilled_attendant, col = continent), size = 1.8) +
  geom_text_repel(aes(haq, skilled_attendant, 
                      label = ifelse(skilled_attendant < 58, country, ''),
                      segment.size = 0.2), 
                  size = 4, max.overlaps = 27) +
  labs(title = "% Births with Skilled Personnel vs. Heathcare Access and Quality Index", 
       x = "HAQ Index", 
       y = "% Births") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()
)

p3 / p4
```

\bigskip

The first scatterplot confirms the relationship the maps were indicating.  The second plot shows how the opposite relationship exists between healthcare quality and the presence of skilled birth attendants.  There is a clear relationship between healthcare access/quality and both 1) factors influencing birth outcomes (% attended by skilled personnel) as well as 2) the outcomes themselves (maternal mortality).

The colors and select labels also corroborate the geographical trends illustrated by the maps.  Countries in Africa and Asia (looking at the map, this is narrowed to South/Southeast Asia) tend to have lower HAQ Indexes, lower proportions of births with skilled personnel, and higher maternal mortality rates.


## Social Attitudes Towards not Having Children

### The Survey Data

The following graphs visualize data from a survey which tracked how much women agreed with certain preconceptions about birth as well as to what extent they felt certain factors influenced the decision to have or not have children.  

```{r, warning = FALSE, message = FALSE}
nj <- read_csv("njsurvey.csv")
q <- read_csv("questions.csv")

nj <- nj %>%
  subset(Gender != "Male")
nj <- subset (nj, select = -Q7) %>%
  rename(employ = "Employment Status")
nj$`Race/Ethnicity`[nj$`Race/Ethnicity` %in% c("Hispanic", "South American- Brazil")] <- "Latinx"

financial <- nj %>%
  subset(select = -c(Q8:Q30)) %>%
  rename(race = "Race/Ethnicity")
financial <- financial %>% pivot_longer(
    cols = starts_with("Q"),
    names_to = "question",
    values_to = "extent",
    values_drop_na = TRUE
  )

society <- nj %>%
  subset(select = -c(Q1:Q6, Q13:Q30)) %>%
  rename(race = "Race/Ethnicity")
society <- society %>% pivot_longer(
    cols = starts_with("Q"),
    names_to = "question",
    values_to = "extent",
    values_drop_na = TRUE
  )

outinfluences <- nj %>%
  subset(select = -c(Q1:Q12, Q21:Q30)) %>%
  rename(race = "Race/Ethnicity")
outinfluences <- outinfluences %>% pivot_longer(
    cols = starts_with("Q"),
    names_to = "question",
    values_to = "extent",
    values_drop_na = TRUE
  )

health <- nj %>%
  subset(select = -c(Q1:Q20)) %>%
  rename(race = "Race/Ethnicity")
health <- health %>% pivot_longer(
    cols = starts_with("Q"),
    names_to = "question",
    values_to = "extent",
    values_drop_na = TRUE
  )
```

```{r, fig.width = 7, fig.height = 6, warning = FALSE, message = FALSE}
question_nums <- unique(society$question)[-4]
question_text <- c("It is reasonable for a woman to not have a child", "Society has a negative perception of women who choose not to have a child", "Women who choose to not have a child are making a poor decision", "Women who choose not to have a child are selfish")

society$question_text <- plyr::mapvalues(society$question, 
          from = question_nums, 
          to = question_text)

society <- society %>% 
  filter(question %in% question_nums) %>%
  group_by(question_text) %>% 
  mutate(avg_extent = mean(extent))

p1 <- ggplot(society, aes(x = extent, y= reorder(question_text, extent, mean), fill = avg_extent)) +
  geom_violin() +
  labs(title = "Societal Preconceptions Surrounding Birth",
       subtitle = "How much do you disagree or agree with the following statements?", 
       x = "1 = strongly disagree, 5 = strongly agree", 
       y = element_blank(),
       fill = "Average Rating") +
  scale_fill_distiller(palette = "RdPu", direction = 1, limits = c(1, 5)) +
  scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 30)) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text=element_text(size=8),
        legend.title=element_text(size=10),
        axis.text=element_text(size=8),
        axis.title.x=element_text(size=10),
        plot.title=element_text(size=12),
        plot.subtitle=element_text(size=10))

p1
```

\bigskip

The survey questions regarding stereotypes and perceptions about women and motherhood show how respondents feel some sense of larger, societal pressure or duty to have children. Respondents largely agreed that women were selfish for not having kids, though they simultaneously valued a woman's choice to have or not have children as "reasonable." Many complex factors influence these opinions, which are explored in the following figures.

### Factors Influencing the Decision to Have/Not Have Children

\bigskip

```{r, fig.width = 7, fig.height = 7, warning = FALSE, message = FALSE}
question_nums <- unique(outinfluences$question)
question_text <- c("Negative stories about childbirth", "Negative messages about single-parent households", "Family/Religion", "Pressure from peers", "Fear of poor parenting skills", "Societal pressure to have children", "Societal perception that same-sex couples cannot have children", "Traditional view that it is a woman's duty to have children")

outinfluences$question_text <- plyr::mapvalues(outinfluences$question, 
          from = question_nums, 
          to = question_text)

outinfluences <- outinfluences %>% 
  group_by(question_text) %>% 
  mutate(avg_extent = mean(extent))

p2 <- ggplot(outinfluences, aes(x = extent, y= reorder(question_text, extent, mean), fill = avg_extent)) +
  geom_violin() +
  labs(title = "Outside Influences",
       subtitle = "To what extent do these factors influence the decision to have children?", 
       x = "1 = no influence, 5 = strong influence",  
       y = element_blank(), 
       fill = "Average Rating") +
  scale_fill_distiller(palette = "RdPu", direction = 1, limits = c(1, 5)) +
  scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 30)) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text=element_text(size=8),
        legend.title=element_text(size=10),
        axis.text=element_text(size=8),
        axis.title.x=element_text(size=10),
        plot.title=element_text(size=12),
        plot.subtitle=element_text(size=10))

p2
```

\bigskip

Societal pressure is a strong influence while more local, personal influences like peer pressure and negative childbirth stores are not.

\bigskip

```{r, fig.width = 7, fig.height = 6, warning = FALSE, message = FALSE}
question_nums <- unique(health$question)[1:6]
question_text <- c("Infertility due to age", "Complications due to age", "Women without children live longer", "Being physically fit before having children", "Pregnancy can cause high blood pressure, depression, or anxiety", "Health risks outweigh benefits")

health$question_text <- plyr::mapvalues(health$question, 
          from = question_nums, 
          to = question_text)

health <- health %>% 
  filter(question %in% question_nums) %>%
  group_by(question_text) %>% 
  mutate(avg_extent = mean(extent)) 

p3 <- ggplot(health, aes(x = extent, y= reorder(question_text, extent, mean), fill = avg_extent)) +
  geom_violin() +
  labs(title = "Health",
       subtitle = "To what extent do these factors influence the decision to have children?", 
       x = "1 = no influence, 5 = strong influence", 
       y = element_blank(),
       fill = "Average Rating") +
  scale_fill_distiller(palette = "RdPu", direction = 1, limits = c(1, 5)) +
  scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 30)) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text=element_text(size=8),
        legend.title=element_text(size=10),
        axis.text=element_text(size=8),
        axis.title.x=element_text(size=10),
        plot.title=element_text(size=12),
        plot.subtitle=element_text(size=10)) 

p3
```

\bigskip

Respondents largely agreed that potential health risks were not strong influences in their decisions to have or not have children.  It is important to remember that all respondents lived in New Jersey and therefore had access to a certain quality of healthcare that is not available in all areas of the globe.

\bigskip

```{r, fig.width = 7, fig.height = 6, warning = FALSE, message = FALSE}
question_nums <- unique(financial$question)[-3]
question_text <- c("Financial instability", "Cost to raise a child", "Opportunity cost of parenting instead of working", "Fertility treatments/adoption costs if infertile", "No financial support from a partner/spouse")

financial$question_text <- plyr::mapvalues(financial$question, 
          from = question_nums, 
          to = question_text)

financial <- financial %>% 
  group_by(question_text) %>% 
  mutate(avg_extent = mean(extent))

p4 <- ggplot(financial, aes(x = extent, y= reorder(question_text, extent, mean), fill = avg_extent)) +
  geom_violin() +
  labs(title = "Financial",
       subtitle = "To what extent do these factors influence the decision to have children?", 
       x = "1 = no influence, 5 = strong influence",  
       y = element_blank(),
       fill = "Average Rating") +
  scale_fill_distiller(palette = "RdPu", direction = 1, limits = c(1, 5)) +
  scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 30)) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text=element_text(size=8),
        legend.title=element_text(size=10),
        axis.text=element_text(size=8),
        axis.title.x=element_text(size=10),
        plot.title=element_text(size=12),
        plot.subtitle=element_text(size=10))

p4
```

\bigskip

Financial factors had very high variance in responses.  This is likely due to differences in respondents’ socioeconomic status.

\bigskip

```{r fig.height = 4, fig.width = 7}
race_financial <- financial %>%
  group_by(race) %>%
  summarise(mean(extent))%>%
  rename("Financial" = "mean(extent)")

race_outinfluences <- outinfluences %>%
  group_by(race) %>%
  summarise(mean(extent)) %>%
  rename("Outside Influences" = "mean(extent)")

race_health <- health %>%
  group_by(race) %>%
  summarise(mean(extent)) %>%
  rename("Health" = "mean(extent)")

df_list <- list(race_financial, race_outinfluences, race_health)

race_factors <- df_list %>% reduce(full_join, by='race')

race_factors <- pivot_longer(race_factors, cols = !race, names_to = "factor", values_to = "extent")

ggplot(race_factors, aes(x= factor, y = extent)) +
  geom_bar(aes(fill = race), stat = "identity",position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Average Extent by Ethnicity", 
       x = element_blank(), 
       y = "Average Extent") +
  guides(fill=guide_legend(title="Ethnicity"))+
  theme_minimal()
```
\bigskip

```{r fig.height = 4, fig.width = 7}
employ_financial <- financial %>%
  group_by(employ) %>%
  summarise(mean(extent))%>%
  rename("Financial" = "mean(extent)")

employ_outinfluences <- outinfluences %>%
  group_by(employ) %>%
  summarise(mean(extent)) %>%
  rename("Outside Influences" = "mean(extent)")

employ_health <- health %>%
  group_by(employ) %>%
  summarise(mean(extent)) %>%
  rename("Health" = "mean(extent)")

df_list <- list(employ_financial, employ_outinfluences, employ_health)

employ_factors <- df_list %>% reduce(full_join, by='employ')

employ_factors <- pivot_longer(employ_factors, cols = !employ, names_to = "factor", values_to = "extent")


ggplot(employ_factors, aes(x= factor, y = extent)) +
  geom_bar(aes(fill = employ), stat = "identity",position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Average Extent by Employment Status", 
       x = element_blank(), 
       y = "Average Extent") +
  guides(fill=guide_legend(title="Employment Status"))+
  theme_minimal()
```

\bigskip


It is important to consider how racial and socioeconomic background (based on employment status) may have played a role in how respondents answered these questions.

\bigskip
\bigskip
\bigskip
\bigskip

## Health Risks Associated with Pregnancy

```{r, warning = FALSE, message = FALSE}
maternal_risk <- read_csv("MaternalHealthRiskDataSet.csv", show_col_types = FALSE)

# update risk factor levels
risks <- (c("low risk", "mid risk", "high risk"))
maternal_risk <- maternal_risk %>%
  mutate(RiskLevel = factor(RiskLevel, levels = risks)) %>%
  filter(HeartRate > 10) # remove outlier with HeartRate of 7
```


We examine how each factor is correlated in the matrix below.

```{r, warning = FALSE, message = FALSE}
mat <- maternal_risk %>%
  mutate(RiskLevel = case_when(RiskLevel == "low risk" ~ 0,
                          RiskLevel == "mid risk" ~ 0.5,
                          RiskLevel == "high risk" ~ 1))
cor(mat)
```

\bigskip

Since we are exploring what may influence maternal risk levels, we choose the three factors with the strongest relationships with risk as indicated by the correlation matrix: blood sugar, systolic and diastolic blood pressure, and age. We explore each of these relationships below.


To examine blood sugar values associated with the three levels of risk, we create a box plot.

\bigskip

```{r fig.height = 3.5, fig.width = 6, warning = FALSE, message = FALSE}
blood_sugar <- maternal_risk %>%
  ggplot() +
  geom_boxplot(aes(BS, RiskLevel, col = RiskLevel)) +
  scale_color_manual(values = c("#eb78c6", "#f0754d", "#40b88e")) +
  labs(title = "Blood Sugar Levels vs. Pregnancy Risk", x = "Blood Sugar") +
  theme_minimal() +
  theme(legend.position = "none") 

blood_sugar
```

\bigskip

It is clear that higher blood sugar levels are correlated with higher pregnancy risk levels, as the high risk box plot has a much larger average and variance. This does not necessarily indicate cause, but does confirm the relatively large correlation value between blood sugar and risk level (0.47329943) seen in the correlation matrix.


Next, we examine blood pressure (both Systolic and Diastolic) and risk.

\bigskip

```{r fig.height = 3.5, fig.width = 6, warning = FALSE, message = FALSE}
blood_pressure <- maternal_risk %>%
  ggplot(aes(DiastolicBP,SystolicBP)) +
  geom_smooth(method = "lm", se = FALSE, color = "lightgray", alpha = 0.2) +
  geom_point(aes(x=DiastolicBP, y=SystolicBP, col = RiskLevel)) +
  scale_color_manual(values = c("#eb78c6", "#f0754d", "#40b88e")) +
  labs(title = "Systolic Blood Pressure vs. Diastolic Blood Pressure", subtitle = "Faceted by Risk Level") +
  facet_wrap(~ RiskLevel) +
  theme_minimal()

blood_pressure
```

\bigskip

It is again clear that high values of these indicators correlates with increased pregnancy risk, though this correlation appears to be slightly weaker than blood sugar. This also confirms the direct relationship between Systolic and Diastolic blood pressure levels predicted in the correlation matrix, as indicated by the trend lines.


Finally, we examine age.

\bigskip

```{r fig.height = 3.5, fig.width = 6, warning = FALSE, message = FALSE}
age <- maternal_risk %>%
  ggplot() +
  geom_boxplot(aes(Age, RiskLevel, col = RiskLevel)) +
  scale_color_manual(values = c("#eb78c6", "#f0754d", "#40b88e")) +
  labs(title = "Age vs. Pregnancy Risk Level") +
  theme_minimal() +
  theme(legend.position = "none")

age
```

\bigskip

Age appears to be directly correlated with risk level. This is expected, as it is commonly assumed that older mothers are likely to have riskier pregnancies. There are some major outliers, however, that indicate that age is not the sole cause of increased risk. For example, there are several mothers over the age of 60 who are still classified as low risk. 

It is important to consider how age might correlate with the other factors recorded in this data set, as it likely influences the other factors. In the above correlation matrix, age is positively correlated with every factor except body temperature, which we previously found to be of little influence on risk. Thus, age on its own may not be a cause of higher-risk pregnancies, but rather an older age is often linked with high values of other factors (blood sugar, blood pressure, etc.) that can increase maternal risk.


## Conclusion
There are multiple aspects that influence a woman’s decision to have or not have children. One major factor is the difference in healthcare access. Many women do not have accessible, quality healthcare, making it difficult for them to seek proper medical help when pregnant. Others include perceptions regarding health, financial, and societal factors that influence prospective mothers. Measurable health factors are also a major influence, including high blood pressure and age, which can make pregnancy risky. These could be deterrents for women thinking of having children. 


If given the opportunity to further develop this project, we would like to add survey data from other areas to be more representative and to consider other factors such as pregnancy and birth education or sexual health practices for women in different areas. These visualizations can be an educational tool for women and considering the access that women have to other resources like this can influence decisions about pregnancy and birth.



## References

Mock, Thomas. (2018). Global Mortality. Retrieved March 13, 2022 from 
https://github.com/rfordatascience/tidytuesday/tree/master/data/2018/2018-04-16.

Our World in Data. (2015). Healthcare Access and Quality Index. Retrieved April 10, 2022 from
https://ourworldindata.org/grapher/healthcare-access-and-quality-index. 

RKKaggle. (2021). Survey on Maternity(NJ, USA). Retrieved April 10, 2022 from
https://www.kaggle.com/datasets/rkkaggle2/social-attitudes-regarding-childlessness-nj-survey?sel
ect=ChildlessnessQuestions.csv

Safrit, Catherine. (2021, December). Maternal Health Risk Data. Retrieved February 
12, 2022 from https://www.kaggle.com/csafrit2/maternal-health-risk-data.

Zeus. (2021). World Health Statistics 2020 | Complete | Geo-Analysis. Retrieved April 10, 2022 
from https://www.kaggle.com/datasets/utkarshxy/who-worldhealth-statistics-2020-complete?select=birthAttendedBySkilledPersonal.csv. 



