---
title: "Milestone2"
author: "Mikayla Lamping"
date: "3/13/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(patchwork)
```

# Maternal Risk

## Check for potential data quality issues

```{r}
maternal_risk <- read_csv("MaternalHealthRiskDataSet.csv")
```

```{r}
# No missing values
sum(maternal_risk %>%
  is.na())
```

Variable definitions from Kaggle:

- Age: Age in years when a woman is pregnant.

- SystolicBP: Upper value of Blood Pressure in mmHg, another significant attribute during pregnancy.

- DiastolicBP: Lower value of Blood Pressure in mmHg, another significant attribute during pregnancy.

- BS: Blood glucose levels is in terms of a molar concentration, mmol/L.

- HeartRate: A normal resting heart rate in beats per minute.

- Risk Level: Predicted Risk Intensity Level during pregnancy considering the previous attribute.


```{r}
summary(maternal_risk)
```

HeartRate seems to have at least one outlier, as the minimum is only 7 but the majority of the points are in the 70-80 bpm range.

### Check for duplicate entries
```{r}
maternal_risk %>%
  distinct()
```

There appear to be many duplicates, as distinct() reduced the number of rows from 1014 to just 452. It does seem that the duplicate entries still represent different people and they just happened to have the same measurements, so these entries should still be included as we proceed with our analysis.



## Build intuition about individual variables

```{r}
histogram <- function(colname) {
  maternal_risk %>%
    ggplot() +
    geom_histogram(aes(x = .data[[colname]]))
}
```

### Age
```{r}
histogram("Age")
```

Age ranges from 10 to 70, with the majority between 15 and 25, which is expected.

### SystolicBP
```{r}
histogram("SystolicBP")
```

Small number of entries with SystolicBP above 150. What are their risk levels?
```{r}
maternal_risk %>%
  filter(SystolicBP > 150)
```

All ten rows are identical and classify each as high risk, which is expected given that these entries are outliers.

### DiastolicBP
```{r}
histogram("DiastolicBP")
```

Fairly evenly distributed. Check under 50 group.
```{r}
maternal_risk %>%
  filter(DiastolicBP < 50)
```

25 entries with Diastolic blood pressure less than 50 (all 49). All are classified as low risk, so upon initial exploration, these low DiastolicBP outliers don't seem to increase risk level.

### Blood Sugar
```{r}
histogram("BS")
```

The majority are under 10, but there are many entries spread between 10 and 20. Do these have higher risk level?
```{r}
maternal_risk %>%
  filter(BS > 10) %>%
  group_by(RiskLevel) %>%
  summarise(count = n())
```

Yes, the majority of the entries with a blood sugar level greater than 10 are classified as high risk. This is a factor we should explore further, as this seems to indicate that high values of BP correlate with higher risk level.

### Body Temperature
```{r}
histogram("BodyTemp")
```

The majority are in the normal temperature range (~98-99).
```{r}
maternal_risk %>%
  filter(BodyTemp > 100) %>%
  group_by(RiskLevel) %>%
  summarise(count = n())
```

There doesn't seem to be a super strong correlation, but there are more mid to high risk entries than low risk in the group of higher body temperatures.


### HeartRate
```{r}
histogram("HeartRate")
```

There is a major outlier, with heart rate less than 10.
```{r}
maternal_risk %>%
  filter(HeartRate < 25)
```
HeartRate may have been entered incorrectly? This person is classified as low risk but has a heart rate of just 7, while the majority of heart rates in this dataset are around 70 bpm.


### Risk Level
```{r}
# Plot discrete histogram (RiskLevel)
maternal_risk %>%
  ggplot() +
  geom_histogram(aes(x = RiskLevel), stat = "count")
```

Risk level mainly low but fairly evenly dispersed.
```{r}
maternal_risk %>%
  count(RiskLevel)
```

The order of the levels isn't intuitive - modify this in next step.


## Prepare processed data
RiskLevel in order that doesn't make sense for plotting. Update factor to have levels increase from low to high.
```{r}
risks <- (c("low risk", "mid risk", "high risk"))

maternal_risk2 <- maternal_risk %>%
  mutate(RiskLevel = factor(RiskLevel, levels = risks))
```

\bigskip

# Global Mortality

## Check for potential data quality issues

```{r}
global <- read_csv("global_mortality.csv")
```

```{r}
# no missing values in data we care about
sum(global %>%
      select(c(`Neonatal deaths (%)`, `Maternal deaths (%)`)) %>%
      is.na())
```
Key Variables:
- Maternal deaths (%): Percentage of all deaths in a given country & year where the cause was linked to maternal/childbirth complications
- Neonatal deaths (%): Percentage of all deaths in a given country & year where death occurred within the first 28 days of life

```{r}
summary(global[c(1, 10, 25)])
```

### Check for duplicate entries
```{r}
# same size as original dataset, no duplicates
nrow(global %>% distinct())
length(global %>% distinct())
```

## Prepare processed data

```{r}
# keep most recent year
global <- global %>%
  filter(year == 2016)
```

## Build intuition about individual variables

```{r fig.height = 24, fig.width = 12}
global <- read_csv("global_mortality.csv") 

countries1 <- ggplot(global) +
  geom_col(aes(`Neonatal deaths (%)`, reorder(country, `Neonatal deaths (%)`), fill = country)) +
  ylab(label = "") +
  theme(axis.text = element_text(size = 2)) +
  theme_minimal() + 
  guides(fill="none")

countries2 <- ggplot(global) + 
  geom_col(aes(`Maternal deaths (%)`, reorder(country, `Maternal deaths (%)`), fill = country)) +
  ylab(label = "") +
  theme(axis.text = element_text(size = 2)) +
  theme_minimal() + 
  guides(fill="none")

countries1 + countries2
```

\bigskip

# Interventions & Maternal Outcomes

## Check for potential data quality issues

The following .csv files were hand-generated from the cited paper (more details in the write-up).

```{r}
interventions <- read_csv("interventions.csv")
outcomes <- read_csv("maternal_outcomes.csv")
```

```{r}
# no missing values
sum(interventions %>%
      is.na())
```
```{r}
sum(outcomes %>%
      is.na())
```

"interventions" Variables:

Number of pregnant people who had...

- external_tocomoter: external electronic fetal monitoring

- fetal_scalp_electrode: internal electronic fetal monitoring

- amniotomy: purposeful puncturing of the amniotic sac 

- oxytocin: drug used to induce labor, brand name Pitocin

- nitrous_oxide: "laughing gas"

- epidural: numbing prodeduce for the lower part of the body

- narcotic: pain-relieving drugs

- spontaneous_vaginal: vaginal birth without forceps or other mechanical interventions

- assisted_vaginal: vaginal birth with mechanical assistance

- cesarean: birth via cesarean-section surgery

- episiotomy: cutting of the perineum for vaginal birth

total: total number in each birth setting/caregiver (same as below)


"outcomes" variables:

Number of pregnant people who experienced...

- prolapsed_cord: a prolapsed umbilical cord

- uterine_rupture: a ruputure of the uterus

- postpartum_hemorrhage: hemorrhaging after giving birth

- blood_transfusion: requiring a blood transfusion

- obstetric_shock: shock, usually caused by excessive blood loss

- death	

- placenta_removal: manual removal of the placenta (rather than natural expulsion)	

- uterine_prolapse: a prolapsed uterus

- pyrexia: fever

- uti: urinary tract infection

- puerpural_fever: prolonged fever immediately after giving birth

- wound_infection: infection at the site of a wound	

- no_tear: no tearing from birth	

- first_second_degree: first or second degree tears

- third_fourth_degree: third or fourth degree tears

- unknown_degree: tearing of unknown degree

- cervical_tear: tearing of the cervix

total: total number in each birth setting/caregiver (same as above)


```{r}
summary(interventions)
```

```{r}
summary(outcomes)
```

Some variables have a very low sample size/count, which is not as useful for visualization.

### Check for duplicate entries

As these two .csv's were made by hand (and are very small overall), we verify there are no duplicate entries.

## Prepare processed data

```{r}
interventions_ratios <- interventions %>%
  mutate_at(vars(-c(setting, caregiver, total)), funs(./total)) %>%
  unite("birth_plan", c(setting, caregiver), sep = ", ")

outcomes_ratios <- outcomes %>%
  mutate_at(vars(-c(setting, caregiver, total)), funs(./total)) %>%
  unite("birth_plan", c(setting, caregiver), sep = ", ")
```

## Build intuition about individual variables

```{r fig.height = 6, fig.width = 8}
i1 <- ggplot(interventions_ratios) +
  geom_col(aes(birth_plan, cesarean, fill = birth_plan)) +
  theme(axis.text = element_text(size = 1)) +
  theme_minimal() +
  guides(fill="none") +
  xlab("")

i2 <- ggplot(interventions_ratios) +
  geom_col(aes(birth_plan, oxytocin, fill = birth_plan)) +
  theme(axis.text = element_text(size = 1)) +
  theme_minimal() + 
  guides(fill="none") +
  xlab("")

i3 <- ggplot(interventions_ratios) +
  geom_col(aes(birth_plan, epidural, fill = birth_plan)) +
  theme(axis.text = element_text(size = 1)) +
  theme_minimal() + 
  guides(fill="none") +
  xlab("")

i4 <- ggplot(interventions_ratios) +
  geom_col(aes(birth_plan, episiotomy, fill = birth_plan)) +
  theme(axis.text = element_text(size = 1)) +
  theme_minimal() + 
  guides(fill="none") +
  xlab("")

((i1 + i2) / (i3 + i4)) + 
    plot_annotation(title = "Interventions")
```

```{r fig.height = 6, fig.width = 8}
o1 <- ggplot(outcomes_ratios) +
  geom_col(aes(birth_plan, blood_transfusion, fill = birth_plan)) +
  theme(axis.text = element_text(size = 1)) +
  theme_minimal() +
  guides(fill="none") +
  xlab("")

o2 <- ggplot(outcomes_ratios) +
  geom_col(aes(birth_plan, pyrexia, fill = birth_plan)) +
  theme(axis.text = element_text(size = 1)) +
  theme_minimal() + 
  guides(fill="none") +
  xlab("")

o3 <- ggplot(outcomes_ratios) +
  geom_col(aes(birth_plan, no_tear, fill = birth_plan)) +
  theme(axis.text = element_text(size = 1)) +
  theme_minimal() + 
  guides(fill="none") +
  xlab("")

o4 <- ggplot(outcomes_ratios) +
  geom_col(aes(birth_plan, third_fourth_degree, fill = birth_plan)) +
  theme(axis.text = element_text(size = 1)) +
  theme_minimal() + 
  guides(fill="none") +
  xlab("")

((o1 + o2) / (o3 + o4)) + 
    plot_annotation(title = "Outcomes")
```